Functors \\ Maps between Categories

### 2017-07-28 notes/notes-2/notes-2.tex
Make some stylistic changes to notes for talk 2

\begin{document}
\title{Maps between Categories -- Functors}
\author{Felix Rech \and Dominik Wagner}
\maketitle

@@

\begin{definition}
  Let $\cat A$ and $\cat B$ be categories. A \demph{functor} $F \from \cat A\to\cat B$ consists of
    \begin{itemize}
    \item a function
      \begin{align*}
        \ob(\cat A)\to\ob(\cat B),
      \end{align*}
      written as $A\mapsto F(A)$ and
    \item for each $A,A'\in\ob(\cat A)$, a function
      \begin{align*}
        \cat A(A,A')\to\cat B(F(A),F(A')),
      \end{align*}
      written as $f\mapsto F(f)$
    \end{itemize}
    such that the following axioms are satisfies:
    \begin{itemize}
    \item $F(1_A)=1_{F(A)}$ for all objects $A$ and
    \item $F(f'\of f)=F(f')\of F(f)$ for all morphisms $A\xrightarrow{f}A'\xrightarrow{f'}A''$.
    \end{itemize}
\end{definition}

The image of such a functor can be thought of as a \textit{shadow} of category $\cat{A}$ in $\cat{B}$.
Sometimes we call the first function \demph{action on paths} and the second function \demph{action on morphisms}.

We can easily define composition of functors and an identity functor on every category.
With those we define a category $\Cat$ that contains categories as objects and functors as morphisms.

@@

You might already spot some similarity to the functoriality axioms.
We use this in our next example.

@@

\begin{example}
  Monoids together with monoid homomorphisms form a category.
  We can define a functor from the category of monoids to $\Set$, that maps every monoid to its element set and every homomorphism to the undelying function.
  This functor maps into a category that is similar to its domain but with less structure on the objects.
  Hence one can say that the functor \emph{forgets} the monoid structure.
  This can be observed on many functors.
  We call such functors \demph{forgetful}.
  Forgetfulness is an informal concept and depends on our interpretation of the category, but usually it is obvious if you should call a functor forgetful.
\end{example}

\begin{example}
  There is a functor from the category of commutative monoids to the category of general monoids which maps every object and every morphism to itself.
  It is injective on objects and on arrows, which means that we can always say where a resulting object \textit{came from}.
  But we still call it \emph{forgetful} because it maps into a category where in general the objects have less structure.
\end{example}

@@

\begin{example}
  Given a set $A$, we define the \demph{free monoid} $A^*$ over $A$ as the set of lists with elements in $A$.
  As binary operation we use the concatenation of lists and as neutral element the empty list.
  If we have a function $A \xrightarrow{f} B$ then we can lift it to a monoid homomorphism $A^* \xrightarrow{f^*} B^*$ that applies $f$ pointwise on a list, i.e.\ $f^*([x, y, \ldots]) = [f(x), f(y), \ldots]$.
  This gives us a functor that maps a set to its free monoid and hence \emph{adds structure}.
  In general we call such a functor a \demph{free functor}.
\end{example}

@@

Note that this definition only speaks about what happens to morphisms between the \emph{same pair} of objects.
For morphisms $A \xrightarrow{f, g} B$, we can conclude that $F(A) = F(B)$ implies $A = B$ but for a morphism $A \xrightarrow{f} B$ and a morphism $C \xrightarrow{g} D$ between another pair of objects, the faithfulness of $F$ does not tell us anything.
Let us consider a functor like in the following graphic as an example.
\[ \begin{tikzpicture}
    \draw (0, 0) rectangle (2, 3);
    \node (A) at (0.5, 2.5) {$A$};
    \node (C) at (1, 0.5) {$C$};
    \node (B) at (1.5, 2.5) {$B$};
    \draw[->] (A) to node[auto, swap] {$f$} (C);
    \draw[->] (B) to node[auto] {$g$} (C);

@@

With our definition, a subcategory is full if and only if the inclusion functor is full.
An important property is that a full subcategory is completely determined by the selection of objects that it contains.

As the image of a monoid homomorphism is a monoid and the image of a group homomorphism is a group, one could easily assume that the image of a functor is a category, but in general this is wrong as we can see in this example:
\[ \begin{tikzpicture}
  \draw (0, 0) rectangle (3, 5);
  \node (A) at (1, 4.5) {$A$};
  \node (B) at (1, 2.5) {$B$};
  \node (C) at (2, 2.5) {$C$};
  \node (D) at (2, 0.5) {$D$};
  \draw[->] (A) to node[auto] {$f$} (B);
  \draw[->] (C) to node[auto] {$g$} (D);

@@

  \draw[->, dotted, in=120, out=45] (A) to (X);
  \draw[->, dotted, in=120, out=45] (B) to (Y);
  \draw[->, dotted, in=-120, out=-45] (C) to (Y);
  \draw[->, dotted, in=-120, out=-45] (D) to (Z);
\end{tikzpicture} \]
Assume that a functor $F$ is given by the dotted arrows and maps the morphism $f$ to $u$ and $g$ to $v$.
Since the image of $F$ contains $X \xrightarrow{u} Y$ and $Y \xrightarrow{i} Z$, it should also contain $v \of u$ to be a category.
But $F$ does not map any morphism to $v \of u$, thus the image is not a category.

>>>>

\begin{document}
\title{Functors \\ Maps between Categories}
\author{Felix Rech \and Dominik Wagner}
\maketitle

@@

\begin{definition}
  Let $\cat A$ and $\cat B$ be categories. A \demph{functor} $F \from \cat A\to\cat B$ consists of
    \begin{itemize}
    \item a function
      \begin{align*}
        \ob(\cat A)\to\ob(\cat B),
      \end{align*}
      written as $A\mapsto F(A)$ and
    \item for all objects $A$ and $A'$, a function
      \begin{align*}
        \cat A(A,A')\to\cat B(F(A),F(A')),
      \end{align*}
      written as $f\mapsto F(f)$
    \end{itemize}
    such that the following axioms are satisfies:
    \begin{itemize}
    \item $F(1_A)=1_{F(A)}$ for all objects $A$ and
    \item $F(f'\of f)=F(f')\of F(f)$ for all morphisms $A\xrightarrow{f}A'\xrightarrow{f'}A''$.
    \end{itemize}
\end{definition}

The image of such a functor can be thought of as a \textit{shadow} of category $\cat{A}$ in $\cat{B}$.
Sometimes we call the first function \demph{action on paths} and the second function \demph{action on morphisms}.
The definition of functors turns the class of categories into a category itself.

\begin{definition}
  We define a category $\Cat$ with categories as elements and functors as morphisms.
  The identity morphism on an object is the functor that maps every object and every morphism to itself.
  The composition is defined by composition of both actions on objects and composition of both actions on morphisms.
\end{definition}

@@

You might already spot some similarity to the functoriality axioms.
We capture this in our next example.

@@

\begin{example}
  Monoids together with monoid homomorphisms form a category.
  We can define a functor from the category of monoids to $\Set$ that maps every monoid to its element set and every homomorphism to the undelying function.
  This functor maps into a category that is similar to its domain but with less structure on the objects.
  Hence one can say that the functor \emph{forgets} the monoid structure.
  This can be observed on many other functors as well.
  We call such functors \demph{forgetful}.
  Forgetfulness is an informal concept and depends on our interpretation of the category, but usually it is obvious if you should call a functor forgetful.
\end{example}

\begin{example}
  There is a functor from the category of commutative monoids to the category of general monoids which maps every object and every morphism to itself.
  It is injective on objects and on arrows, which means that we can always say, where a resulting object \textit{came from}.
  But we still call it \emph{forgetful} because it maps into a category where in general the objects have less structure.
\end{example}

@@

\begin{example}
  Given a set $A$, we define the \demph{free monoid} $A^*$ over $A$ as the set of lists with elements in $A$.
  As binary operation we use the concatenation of lists and as neutral element the empty list.
  If we have a function $A \xrightarrow{f} B$ between sets then we can lift it to a monoid homomorphism $A^* \xrightarrow{f^*} B^*$ that applies $f$ pointwise on a list, i.e.\ $f^*([x, y, \ldots]) = [f(x), f(y), \ldots]$.
  This gives us a functor that maps a set to its free monoid and hence \emph{adds structure}.
  In general we call such a functor a \demph{free functor}.
\end{example}

@@

Note that this definition only speaks about what happens to morphisms between the \emph{same pair} of objects.
We can still map two different morphisms onto the same target, as long as they are defined between a different pair of objects.
Let us illustrate this with an example.
\[ \begin{tikzpicture}
    \draw (0, 0) rectangle (2, 3);
    \node (A) at (0.5, 2.5) {$A$};
    \node (C) at (1, 0.5) {$C$};
    \node (B) at (1.5, 2.5) {$B$};
    \draw[->] (A) to node[auto, swap] {$f$} (C);
    \draw[->] (B) to node[auto] {$g$} (C);

@@

With our definition, a subcategory is full if and only if the inclusion functor is full.
A convenient property is that a full subcategory is completely determined by the selection of objects that it contains.

As the image of a monoid homomorphism is a monoid and the image of a group homomorphism is a group, one could easily assume that the image of a functor is a category, but in general this is wrong as we can see in the next example:
\[ \begin{tikzpicture}
  \draw (0, 0) rectangle (3, 5);
  \node (A) at (1, 4.5) {$A$};
  \node (B) at (1, 2.5) {$B$};
  \node (C) at (2, 2.5) {$C$};
  \node (D) at (2, 0.5) {$D$};
  \draw[->] (A) to node[auto] {$f$} (B);
  \draw[->] (C) to node[auto] {$g$} (D);

@@

  \draw[->, dotted, in=120, out=45] (A) to (X);
  \draw[->, dotted, in=120, out=45] (B) to (Y);
  \draw[->, dotted, in=-120, out=-45] (C) to (Y);
  \draw[->, dotted, in=-120, out=-45] (D) to (Z);
\end{tikzpicture} \]
Assume that a functor $F$ is given by the dotted arrows and maps the morphism $f$ to $u$ and $g$ to $v$.
Since the image of $F$ contains $u$ and $v$, it should also contain $v \of u$ to be a category.
But $F$ does not map any morphism to $v \of u$, thus the image is not a category.
